Mechanising Recursion Schemes with Magic-Free Coq Extraction

### 2024-10-08 paper/paper.tex
minor
Stuff

\section{Introduction}
\label{sec:intro}
Structured recursion schemes~\cite{HinzeW16, HinzeWG15} are powerful
abstractions that capture common patterns of recursion. The main benefit of
structuring computation using recursion schemes, is that they enjoy
well-estabished \emph{algebraic properties} that can serve as a foundation for
program optimisation (e.g.\ \emph{fusion} laws, or semi-automatic
parallelisations~\cite{TakanoM95,Gibbons96:Third, farmsCastro}).  These
algebraic properties led to their use in the context of \emph{program
calculation}, where programmers would describe their code using simple,
inefficient specifications in an \emph{algebra of programming}~\cite{}, and then
use the algebraic laws of this algebra of programming to \emph{calculate} an
efficient version of the same algorithm.

\dcas{TODO: Introduce example}

Despite the rigorous algebraic laws that are satisfied by recursion schemes,
there is a lack of tool support for their use in the context of program
calculation. In fact, most of the work on applying program calculation is done
by performing pen-and-paper proofs, and then translating the result to specific
instances of recursion schemes, implemented in a programming language (generally
Haskell)~\cite{TODO:INeedToFindSomeMoreReferences}. Most of the work on
automating program calculation techniques either build implementations of the
algebraic laws, together with some form of syntax or
EDSL~\cite{TODO:INeedToFindSomeMoreReferences}. Such implementations are not
verified, and can introduce bugs in the way they apply the algebraic laws as
rewriting rules.

A more rigorous approach to implementing program calculation techniques is to
\emph{mechanise} them as part of a proof assistant. However, most of the work in
mechanising recursion schemes focuses on a narrow subset of the known recursion
schemes due to termination
issues~\cite{TODO:FillInReferencesForCoqMechanisationsFromRelatedWorkSection}.
Indeed, most of the mechanisations focus on \emph{maps} and \emph{folds}, and
few authors focus on the \emph{unfolds}. In general, Hinze et
al.~\cite{HinzeWG15} has proven that all known  recursion schemes are instances
of a  divide-and-conquer algorithm which goes under the name of
\emph{hylomorphism}~\cite{MeijerFP91,HuIT96}.  To date, no work has mechanised
hylomorphisms in the Coq proof assistant.

@@

Recently, Abreu et al.~\cite{AbreuDHJMS23} encoded an algebraic approach to
divide-and-conquer computations in which termination is entirely enforced by the
typing discipline. Their approach solves the problem of termination proofs as
well as the performance of the code that is run \emph{within Coq}, but it does
not allow for extraction of idiomatic OCaml code. This is unfortunate since code
extration is what allows the execution of code that has been verified in
Coq~\cite{OnoHTNH11, Larchey-Wendling23, MiculanP12, Sakaguchi20}.  In Abreu et
al.'s approach, extraction (1) do not preserve the recursive structure of common
implementations; and (2) lead to unsafe casts like \ocaml{Obj.magic} in the
generated code. This latter is also problematic in that, for higher-order
programs, simple interoperations can lead to incorrect behaviour or even
segfaults~\cite{forster:hal-04329663} and, moreover, it invalidates the
fast-and-loose principle~\cite{DanielssonHJG06}.

This work presents the first Coq formalisation \emph{hylomorphisms} that (1) is
\emph{axiom-free} and (2) allows the extraction of idiomatic OCaml code. The
full mechanisation can be found on
Github\footnote{https://github.com/dcastrop/coq-hylomorphisms}. While
programmers still need to reason about the termination of their programs, our
mechanisation will allow the use of the algebraic laws of hylomorphisms for
\emph{program calculation}, as well as extracting to idiomatic code (see
extracted \ocaml{qsort} program in Section~\ref{sec:examples}).
% As an example, our mechanisation of \ocaml{qsort} as a
% hylomorphism will produce the following code:
% \begin{coqcode}
% let rec qsort = function | [] -> []
%   | h :: t -> let (l, r) = partition (fun x0 -> (<=) x0 h) t in
%               let cx = fun p -> match p with | Lbranch -> l | Rbranch -> r in
%               app (qsort (cx Lbranch)) (h :: (qsort (cx Rbranch)))
% \end{coqcode}

>>>>

\section{Introduction}
\label{sec:intro}
Structured recursion schemes~\cite{HinzeW16, HinzeWG15} are powerful
abstractions that capture common patterns of recursion. The main benefit of
structuring computation using recursion schemes, is that they enjoy
well-estabished \emph{algebraic properties} that can serve as a foundation for
program optimisation (e.g.\ \emph{fusion} laws, or semi-automatic
parallelisations~\cite{TakanoM95,Gibbons96:Third, Morihata09:Third,
farmsCastro}).  These algebraic properties led to their use in the context of
\emph{program calculation}, where programmers would describe their code using
simple, inefficient specifications in an \emph{algebra of
programming}~\cite{BirddeMoor96:Algebra}, and then use the algebraic laws of
this algebra of programming to \emph{calculate} an efficient version of the same
algorithm. Suppose, for example, that we want to write a program sorts a list of
integers, and then multiplies by 2 all its elements. In OCaml, we may write this
function directly:
\begin{minted}{ocaml}
let rec sort_times_two = function
  | [] -> []
  | h :: t -> let (l, r) = partition (fun x -> x < h) t in
  sort_times_two l @ (h * 2) :: sort_times_two t
\end{minted}
Instead of writing this function directly, we may observe that we can derive it
by \emph{fusing} a regular \emph{quickort} OCaml implementation, with
\ocaml{map (fun x -> x * 2)}. This is the main idea behind program calculation:
start
from a simple specification, e.g.\
$\mathsf{map}\;(\lambda x.\;x \times 2) \circ \mathsf{sort}$, and use program
equivalences and algebraic laws to rewrite it to an optimised version  (e.g.\
the fused OCaml implementation above). We will revisit a similar example
in Section~\ref{sec:sorting}.

Despite the rigorous set of algebraic laws that are satisfied by recursion
schemes, there is a lack of tool support for their use in the context of program
calculation. In fact, most of the work on applying program calculation is done
by performing pen-and-paper proofs, and then translating the result to specific
instances of recursion schemes, implemented in a programming language (generally
Haskell). There are some examples of implementations of program calculation
techniques, but these implementations are scarce, not up to date, and not
verified. For example, Cunha et al. automated program calculation techniques by
a custom Haskell implementation~\cite{DrHylo}.

Several authors build tools for applying program calculation techniques by
\emph{mechanising} them as part of a proof assistant. However, most of the work
in mechanising recursion schemes focuses on a narrow subset of the known
recursion schemes due to termination issues, or do not focus on proving
algebraic laws of generic recursion schemes~\cite{10.1007/978-3-642-17796-5_10,
MurataE19, larchey2022braga}.  Indeed, most of the mechanisations of generic
recursion schemes focus on \emph{maps} and \emph{folds}, and few authors focus
on the \emph{unfolds}. In general, Hinze et al.~\cite{HinzeWG15} has proven that
all known  recursion schemes are instances of a  divide-and-conquer algorithm
which goes under the name of \emph{hylomorphism}~\cite{MeijerFP91,HuIT96}.  To
date, no work has mechanised hylomorphisms in the Coq proof assistant, together
with their algebraic laws.

@@

Recently, Abreu et al.~\cite{AbreuDHJMS23} encoded an algebraic approach to
divide-and-conquer computations in which termination is entirely enforced by the
typing discipline. Their approach solves the problem of termination proofs as
well as the performance of the code that is run \emph{within Coq}, but it does
not allow for extraction of idiomatic OCaml code. This is unfortunate since code
extration is what allows the execution of code that has been verified in
Coq~\cite{OnoHTNH11, Larchey-Wendling23, MiculanP12, Sakaguchi20}.  In Abreu et
al.'s approach, extraction (1) does not preserve the recursive structure of
common implementations; and (2) leads to unsafe casts like \ocaml{Obj.magic} in
the generated code. This latter is also problematic in that, for higher-order
programs, simple interoperations can lead to incorrect behaviour or even
segfaults~\cite{forster:hal-04329663} and, moreover, it invalidates the
fast-and-loose principle~\cite{DanielssonHJG06}.

This work presents the first Coq formalisation \emph{hylomorphisms} that (1) is
\emph{axiom-free} and (2) allows the extraction of idiomatic OCaml code. The
full mechanisation can be found on
Github\footnote{https://github.com/dcastrop/coq-hylomorphisms}\dcas{I SHOULD
ANONYMISE THIS}. While programmers still need to reason about the termination of
their programs, our mechanisation will allow the use of the algebraic laws of
hylomorphisms for \emph{program calculation}, as well as extracting to idiomatic
code (see examples of extracted code in Section~\ref{sec:examples}).
% As an example, our mechanisation of \ocaml{qsort} as a
% hylomorphism will produce the following code:
% \begin{coqcode}
% let rec qsort = function | [] -> []
%   | h :: t -> let (l, r) = partition (fun x0 -> (<=) x0 h) t in
%               let cx = fun p -> match p with | Lbranch -> l | Rbranch -> r in
%               app (qsort (cx Lbranch)) (h :: (qsort (cx Rbranch)))
% \end{coqcode}
